### 2019-10-19 ThesisJo/2_background.tex
2019-10-18

%&header
% !TeX root = 1_introduction

@@

% Related work can be found e.g. in ultrasound imaging, where Chen et. al have investigated porting a time domain MVDR method to a GPU \cite{Chen2009,Chen2011,Chen2011a}. Unlike their work, however, we do not impose the restriction of the input data to be real, nor do we approach larger systems by enforcing Toeplitz structure of the covariance matrix. 
% \\
\begin{figure}[tp]\label{2_fig_article_relation}
\makebox[\linewidth][c]{%
\graphicsAI[drawing,width=1.3\linewidth]{gfx/article_relation.svg}}
\caption{Method relations. This thesis explores three beamformers reconstructing images from acoustic echoes: time-domain MVDR (reference), frequency-domain MVDR (close approximation) and time-domain LCA (educated trial-and-error). Article IV presents background on synthesis of image templates from reconstructed images. }
\end{figure}

\begin{figure}[!t]\label{2_fig_geometry}\centering
\makebox[\linewidth][c]{%
\graphicsAI[width=.7\linewidth]{gfx/geometry.svg}}
\caption{
An array with $M$ sensors has reference frame $\udot{A} \triangleq (\udot{A},\uvec{A})$ attached to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (\,$\hat{\cdot}$ denoting unit length). Impinging on the array at angle $\theta$ is a monochromatic plane wave: a single frequency wave from a source ${S}$ far away relative to the wavelength and array length (far field). Notation details are presented in Paper IV.
}
\end{figure}

Any wave, regardless of its nature, is governed by a partial differential equation known as the \emph{wave equation}:
%
\begin{align}
\nabla^2 f 
 = \frac{1}{c}\frac{\delta^2 f}{\delta t^2},
\end{align}
%\frac{\delta^2 f}{\delta x^2} + \frac{\delta^2 f}{\delta y^2} + \frac{\delta^2 f}{\delta z^2}
%
where $\nabla^2$ is the Laplace operator, $f$ is field strength and $c$ is wave propagation speed. If $f$ represents the displacement of a vibrating string, the equation states that the force pulling the string towards equilibrium (right side) is proportional to the string's curvature (left side). In acoustics $f$ represents sound pressure, but since the mathematics are independent of the physics involved we simply think of $f$ as a signal: a unit-less scalar function with wave characteristics.

The wave equation has a few solutions of interest, in particular that of a plane and spherical wave:
%
\begin{align}
f_\textrm{plane}(\vec{p},t)  &= A e^{j(\omega t - \vec{k} \cdot \vec{p}_{OS})} \\
f_\textrm{spheric}(\vec{p},t) &= \frac{A}{\lVert \vec{p}_{OS} \rVert} e^{j(\omega t - \lVert\vec{k}\rVert \lVert \vec{p}_{OS} \rVert)},
\end{align}
% 
Here $A$ is amplitude, $\omega$ is temporal wave frequency, $t$ is time, $\lVert\vec{k}\rVert=\frac{\omega}{c}=\frac{2\pi}{\lambda}$ is spatial wave frequency (wavenumber) and $\vec{p}_{OS}$ is position of the source $\mathcal{S}$ relative to the observer $\mathcal{O}$. We model the wavefield as superimposed plane waves under far-field conditions (when distance to the source is large relative to the wavelength and array length) and as superimposed spherical waves under near-field conditions.

@@

\begin{figure}[tp]\label{2_fig_scenario_das}
\makebox[\linewidth][c]{%
\graphicsAI[width=1\linewidth]{gfx/scenario_interference_das.svg}}
\caption{DAS beampattern is static (conventional beamforming). Interference suppression at the first mainlobe is -13\,dBr.}
\end{figure}

\begin{figure}[tp]\label{2_fig_scenario_mvdr}
\makebox[\linewidth][c]{%
\graphicsAI[width=1\linewidth]{gfx/scenario_interference_mvdr.svg}}
\caption{MVDR beampattern is dynamic and continuous (adaptive beamforming). Interference suppression is good for prevalent noise sources if a good estimate of the spatial statistics can be obtained.}
\end{figure}

\begin{figure}[tp]\label{2_fig_scenario_lca}
\makebox[\linewidth][c]{%
\graphicsAI[width=1\linewidth]{gfx/scenario_interference_lca.svg}}
\caption{LCA beampattern is dynamic and discrete (adaptive beamforming). Interference suppression is on-par with MVDR with 15 weight sets to choose from, nearly as good even with 6 weight sets.}
\end{figure}

Imagine an array with $M$ sensors at positions $\udot{0},\udot{1},\dots, \udot{m}, \dots, \udot{M{\!-\!1}}$ with reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ attached to its phase center (Figure \ref{2_fig_geometry}), where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation ( $\hat{\cdot}$ denoting unit length). Assume that the array channels are 
digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n-\Delta_m]\in\mathbb{C}$. Focus at source $\udot{S}$ requires delays akin to
%
\begin{align}
\Delta_m[n] = \frac{\lVert \vec{p}_{m}\rVert - \lVert \vec{p}_{0S}\rVert}{c}.\label{2_eq_delta}
\end{align}
%
where $\vec{p}_{m}$ denotes position vector $\udot{A}$ to $\udot{m}$ and $\vec{p}_{0S}$ denotes position vector $\udot{0}$ to $\udot{S}$, respectively. By making the dependence on $\theta$ implicit and assuming the data to be pre-delayed to the pixel of interest, we simplify the notation to $x_m[n]$. By definition, the beamformer output $z[n]\in\mathbb{C}$ is the weighted sum of delayed data samples:
%
\begin{align}
z[n] = \dvec{w}^H[n]\dvec{x}[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]} \in\mathbb{C},\label{2_eq_z}
\end{align}
%
where $w_m\in\mathbb{C}$ is the weight factor assigned to channel $m$. Spatial response of the weights when the array is steered to broadside can be characterized by the beampattern,
%
\begin{align}
W(\vec{k}) = \sum_m^{M-1} w_m^* e^{j\vec k \cdot \vec{p}_m}.\label{2_eq_beampattern}
\end{align}
%
%
%
%
%. Beampattern is broadside-steered beamformer output in the presence of a single plane wave, as a function of the wave's angle.
%%Steered response is beamformer's output in the presence of some wave field, as a function of steering angle. 
%
The beamformer is called delay-and-sum (DAS) or conventional if the weights are static. A variety of weighting functions exist for trading lateral resolution for improved noise suppression (contrast), but a compromise between the two is unavoidable~\cite{Harris1978}. A class of adaptive beamformers target this limitation by adapting the weights to the impinging wavefield. One such technique is the minimum variace distortionless response (MVDR) method.

@@

%\begin{figure}[t]\centering%
%\graphicsAI[drawing,width=1\linewidth]{gfx/implementation.svg}%
%\caption{MVDR beamforming. For one of the total number of pixels in range and azimuth, $N_y$ and $N_x$,\newline
%1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
%2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
%3. and the beamformer output is computed from $z$ from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{2_eq_weights}). }\label{mvdr_beamforming}
%\end{figure}
%
The basic idea behind MVDR is to compute the set of weights that minimizes noise and interference accumulated by the array, while ensuring unit gain for signals coming from the pixel of interest. With pre-delayed data this can be formulated as follows:
%
\begin{gather}
\argmin{\dvec{w}}\ E\Big\{\big|\dvec{w}^H[n]\dvec{x}[n]\big|^2\Big\} \qquad\text{subject to}\qquad \dvec{w}^H[n]\dvec{1} = 1. \label{2_eq_mvdr_def}
% \vec w[n] = \frac{\Ri[n]\1}{\1^T\Ri[n]\1},\label{2_eq_weights}
\end{gather}
%
This optimization problem has the solution:
%
\begin{gather}
\vec w[n] = \frac{\Ri[n]\dvec{1}}{\dvec{1}^T\Ri[n]\dvec{1}},\label{2_eq_weights}
\end{gather}
%
where $\dvec1$ is a row vector of ones representing broadside steering and $\R=E\{\x[n]\x^H[n]\}$ is the spatial covariance matrix for the full array. We estimate $\R$ by computing a sample covariance matrix $\eR$. A good estimate requires some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Synnevag2009a}. Combined, these techniques will also ensure a numerically well conditioned $\eR$.

To arrive at a formulation of how $\eR$ can be estimated, we first split the array into subarrays of length $L$ that overlap on all but one sensor. If we let $x_l[n]$ represent the datavector from subarray $l$,
%
\begin{gather}
\dvec{x}_l[n] = \bmat{x_l[n] & x_{l+1}[n] & \dots & x_{l+L-1}[n]}^T.
\end{gather}
%
then the sample covariance matrix formed with temporal and spatial averaging, $\eR$, is given as:
%
\begin{gather}
\eR[n] =  \frac{1}{N_K N_L} \sumb{l=0}{N_L-1}\sumb{n'=n-K}{n+K} \x_l[n]\x_l^H[n] \in\mathbb{C}^{L,L},\label{2_eq_spatialR}
\end{gather}
%
where $N_K = 2K+1$ is the number of temporal samples to perform averaging over, and $N_L = M-L+1$ is the number of subarrays. In the actual design a fraction of the total power of $\eR[n]$ is finally added to its diagonal~\cite{Synnevag2007}, but this step will be ignored here due to its negligible contribution to runtime.

>>>>

%&header
% !TeX root = thesis
%% !TeX root = 1_introduction

@@

% Related work can be found e.g. in ultrasound imaging, where Chen et. al have investigated porting a time domain MVDR method to a GPU \cite{Chen2009,Chen2011,Chen2011a}. Unlike their work, however, we do not impose the restriction of the input data to be real, nor do we approach larger systems by enforcing Toeplitz structure of the covariance matrix. 
% \\
\begin{figure}[tp]
\makebox[\linewidth][c]{%
\graphicsAI[drawing,width=1.3\linewidth]{gfx/article_relation.svg}}
\caption{Method relations. This thesis explores three beamformers reconstructing images from acoustic echoes: time-domain MVDR (reference), frequency-domain MVDR (close approximation) and time-domain LCA (educated trial-and-error). Article IV presents background on synthesis of image templates from reconstructed images. }%
\label{2_fig_article_relation}
\end{figure}

\begin{figure}[!t]\centering
\makebox[\linewidth][c]{%
\graphicsAI[width=.7\linewidth]{gfx/geometry.svg}}
\caption{
An array with $M$ sensors has reference frame $\udot{A} \triangleq (\udot{A},\uvec{A})$ attached to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (\,$\hat{\cdot}$ denoting unit length). Impinging on the array at angle $\theta$ is a monochromatic plane wave: a single frequency wave from a source $\mathcal{S}$ far away relative to the wavelength and array length (far field). Notation details are presented in 
}%
\label{2_fig_geometry}
\end{figure}

Any wave, regardless of its nature, is governed by a partial differential equation known as the \emph{wave equation}. For a linear, homogeneous (translation-invariant) and isotropic (rotation-invariant) medium its form is
%
\begin{align}\label{eq_wave_equation}
\nabla^2 f 
 = \frac{1}{c}\frac{\delta^2 f}{\delta t^2},
\end{align}
%\frac{\delta^2 f}{\delta x^2} + \frac{\delta^2 f}{\delta y^2} + \frac{\delta^2 f}{\delta z^2}
%
where $\nabla^2$ is the Laplace operator, $f$ is field strength and $c$ is wave propagation speed. In complex mediums $c$ may depend on signal strength (non-linear), position (non-homogeneous/refractive), angle (anisotropic), frequency (dispersive) or be non-real (lossy). Deep sea is acoustically non-homogeneous due to varying pressure, temperature and salinity, but has negligible dispersion and attenuation. Biological tissue both non-homogeneous, non-linear and lossy. We approximate both cases with (\ref{eq_wave_equation}), because data inaccuracies are inevitable and good for testing algorithmic robustness.

If $f$ represents displacement of a vibrating string, (\ref{eq_wave_equation}) states that the force pulling the string towards equilibrium (right side) is proportional to the string's curvature (left side). In acoustics $f$ represents sound pressure, with SI unit Pascal [Pa = N/m$^2$] or logarithmic unit sound-pressure-level (SPL) [1\,Pa = 94\,$\mathrm{dB}_{\textrm{SPL}}$]. To generalize and simplify, we keep our equations free of physical units and simply think of $f$ as a signal: a unit-less scalar function with wave characteristics.

The wave equation has a few solutions of interest, in particular that of a plane and spherical wave:
%
\begin{align}
f_\textrm{plane}(t,\vec{p}_{OS})  &= A e^{j(\omega t - \vec{k} \cdot \vec{p}_{OS})} \\
f_\textrm{spheric}(t,\vec{p}_{OS}) &= \frac{A}{\lVert \vec{p}_{OS} \rVert} e^{j(\omega t - \lVert\vec{k}\rVert \lVert \vec{p}_{OS} \rVert)},
\end{align}
% 
Here $A$ is amplitude, $\omega$ is temporal wave frequency, $t$ is time, $\lVert\vec{k}\rVert=\frac{\omega}{c}=\frac{2\pi}{\lambda}$ is spatial wave frequency (wavenumber) and $\vec{p}_{OS}$ is position of the source $\mathcal{S}$ relative to the observer $\mathcal{O}$. We model the wavefield as superimposed plane waves under far-field conditions (when distance to the source is large relative to wavelength and array length) and as superimposed spherical waves under near-field conditions.

@@

\begin{figure}[tp]
\makebox[\linewidth][c]{%
\graphicsAI[width=1\linewidth]{gfx/scenario_interference_das.svg}}
\caption{DAS (delay-and-sum) beampattern is static (conventional beamforming). Interference suppression at the first mainlobe is -13\,dBr.}%
\label{2_fig_scenario_das}
\end{figure}

\begin{figure}[tp]
\makebox[\linewidth][c]{%
\graphicsAI[width=1\linewidth]{gfx/scenario_interference_mvdr.svg}}
\caption{MVDR (minimum variance distortionless response) beampattern is dynamic and continuous (adaptive beamforming). Interference suppression is good for prevalent noise sources assuming sufficient spatial statistics.}%
\label{2_fig_scenario_mvdr}
\end{figure}

\begin{figure}[tp]
\makebox[\linewidth][c]{%
\graphicsAI[width=1\linewidth]{gfx/scenario_interference_lca.svg}}
\caption{LCA (low complexity adaptive) beampattern is dynamic and discrete (adaptive beamforming). Interference suppression is on-par with MVDR with 15 weight sets to choose from, marginally worse with 6 weight sets.}%
\label{2_fig_scenario_lca}
\end{figure}

Imagine an array $\mathcal{A}$ with $M$ sensors at positions $\udot{0},\udot{1},\dots, \udot{m}, \dots, \udot{M{\!-\!1}}$ with reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ attached to its phase center (Figure \ref{2_fig_geometry}), where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation ( $\hat{\cdot}$ denoting unit length). Assume the array channels are digitally delayed to focus at $\udot{S}$, such that the delayed data from the $m$th channel at instant $n$ can be expressed as $x_m[n-\Delta_m, \vec{p}_{mS}]\in\mathbb{C}$. Focus at source $\udot{S}$ requires delays akin to
%
\begin{align}
\Delta_m[n] = \frac{\lVert \vec{p}_{mS}\rVert - \underset{m'}{\mathrm{min}}\ \lVert \vec{p}_{m'S}\rVert}{c},\label{2_eq_delta}
\end{align}
%
where both $\vec{p}_{mS}$ and $\vec{p}_{m'S}$ denote position vector $\udot{m}$ to $\udot{S}$ and subtracting the minimum distance is optional. By making the dependence on $\vec{p}$ implicit and assuming the data to be pre-delayed to the pixel of interest, we simplify the notation to $x_m[n]$. By definition, the beamformer output $z[n]\in\mathbb{C}$ is the weighted sum of delayed data samples:
%
\begin{align}
z[n] = \dvec{w}^H[n]\dvec{x}[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]} \in\mathbb{C},\label{2_eq_z}
\end{align}
%
where $w_m\in\mathbb{C}$ is the weight factor assigned to channel $m$. Spatial responses of the weights when the array is steered to broadside can be characterized by the beampattern,
%
\begin{align}
W(\vec{k}) = \sum_m^{M-1} w_m^* e^{j\vec k \cdot \vec{p}_m}.\label{2_eq_beampattern}
\end{align}
%
%
%
%
%. Beampattern is broadside-steered beamformer output in the presence of a single plane wave, as a function of the wave's angle.
%%Steered response is beamformer's output in the presence of some wave field, as a function of steering angle. 
%
The beamformer is called delay-and-sum (DAS) or conventional if the weights are static. A variety of weighting functions exist for trading lateral resolution for improved noise suppression (contrast), but a compromise between the two is unavoidable~\cite{Harris1978}. A class of adaptive beamformers target this limitation by adapting the weights to the impinging wavefield. One such technique is the minimum variance distortionless response (MVDR) or Capon beamformer, named after its maker for use in spectral estimation~\cite{Capon1969}.

% Assume:
% Known array characteristics (sensor locations, element patterns, ...
% Stationary signal and noise
% Long observation times
% Known direction of arrival

@@

%\begin{figure}[t]\centering%
%\graphicsAI[drawing,width=1\linewidth]{gfx/implementation.svg}%
%\caption{MVDR beamforming. For one of the total number of pixels in range and azimuth, $N_y$ and $N_x$,\newline
%1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
%2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
%3. and the beamformer output is computed from $z$ from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{2_eq_weights}). }\label{mvdr_beamforming}
%\end{figure}
%

Time-domain MVDR often improves image contrast and resolution compared to conventional beamformers, as demonstrated in e.g. radar~\cite{Benitz1997}, ultrasound~\cite{Synnevag2007} and active sonar~\cite{Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}. It achieves this by computing the weight set that minimizes noise and interference accumulated by the array, under the constraint of unit gain for signals from the pixel of interest. Assuming pre-delayed data, we can formulate this as
%
\begin{gather}
\argmin{\dvec{w}}\ E\Big\{\big|\dvec{w}^H[n]\dvec{x}[n]\big|^2\Big\} \qquad\text{subject to}\qquad \dvec{w}^H[n]\dvec{a} = 1, \label{2_eq_mvdr_def}
% \vec w[n] = \frac{\Ri[n]\1}{\1^T\Ri[n]\1},\label{2_eq_weights}
\end{gather}
%
where $\dvec{a}$ is steering vector. Applying Lagrange multipliers to this convex optimization problem yields the solution
%
\begin{gather}
\dvec w[n] = \frac{\Ri[n]\dvec{a}}{\dvec{a}^T\Ri[n]\dvec{a}} \qquad\text{where}\qquad \dvec{a}=\dvec{1}, \label{2_eq_weights}
\end{gather}
%
where $\dvec{1}$ is a row-vector of ones representing broadside phase-steering and $\R=E\{\dvec{x}[n]\dvec{x}^H[n]\}$ is the spatial covariance matrix for the full array. (\ref{2_eq_weights}) is the maximum likelihood estimate of the signal in the presence of zero-mean random Gaussian noise. We estimate $\R$ with a sample covariance matrix $\eR$, requiring some degree of
%
\begin{itemize}
\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985};
\item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979,VanTrees2002}.
\end{itemize}%

requiring some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics, and \emph{diagonal loading} to improve robustness to parameter errors~\cite{Synnevag2009a}.

Combined, these techniques also ensure a numerically well conditioned $\eR$.

\begin{itemize}
\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985};
\item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
\end{itemize}%

We split the array into subarrays of length $L$ that overlap on all but one sensor. If we let $x_l[n]$ represent the datavector from subarray $l$,
%
\begin{gather}
\dvec{x}_l[n] = \bmat{x_l[n] & x_{l+1}[n] & \dots & x_{l+L-1}[n]}^T,
\end{gather}
%
then the sample covariance matrix formed with temporal and spatial averaging, $\eR$, is given as
%
\begin{gather}
\eR[n] =  \frac{1}{N_K N_L} \sumb{l=0}{N_L-1}\sumb{n'=n-K}{n+K} \x_l[n]\x_l^H[n] \in\mathbb{C}^{L,L},\label{2_eq_spatialR}
\end{gather}
%
where $N_K = 2K+1$ is the number of temporal samples to perform averaging over, and $N_L = M-L+1$ is the number of subarrays. In the actual design a fraction of the total power of $\eR[n]$ is finally added to its diagonal~\cite{Synnevag2007}, but this step will be ignored here due to its negligible contribution to runtime.

### 2019-10-19 ThesisJo/3_publications.tex
2019-10-18

This article accelerates time-domain MVDR with GPUs for use in active sonar imaging. Time-domain MVDR is the reference method for creating sonar images using the minimum variance optimization, aimed at reducing noise and interference while preserving signal integrity. We refer to it as fully adaptive because it computes array weights analytically from spatial features in the data, yielding array weights from a continuous solution space that are optimal in the minimum variance sense. Like any data-driven algorithm, it requires statistically distinguishable signal and noise. However, in typical active systems, the signal is highly correlated with the noise and has poor temporal sample support. Therefore, in its unconstrained form, MVDR overfits the data, yields high variance and fails to generate representative images.

>>>>

This article accelerates time-domain MVDR with GPUs for use in active sonar imaging. Time-domain MVDR is the reference method for creating sonar images using minimum variance optimization, aimed at reducing noise and interference while preserving signal integrity. We refer to it as fully adaptive because it computes array weights analytically from spatial features in the data, yielding array weights from a continuous solution space that are optimal in the minimum variance sense. Like any data-driven algorithm, it requires statistically distinguishable signal and noise. However, in typical active systems, the signal is highly correlated with the noise and has poor temporal sample support. Therefore, in its unconstrained form, MVDR overfits the data, yields high variance and fails to generate representative images.
